---
title: "CodeBook"
author: "Paulo Medina"
date: "2025-02-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Directory and workplace prep

I loaded the necessary packages as well as created a data directory to store the tables created.

```{r}
# Loading necessary packages

#install.packages("dplyr")
#install.packages("here")
#install.packages("plyr")
library(plyr)
library(dplyr)
library(here)

#Set-up directory for work and data

if(!dir.exists("./data")){dir.create(here("./data"))}

```


## Data extraction

In order to retrieve the data, I loaded and saved the data to a temporary variable "Temp". After that, I extracted all the tables with data.

```{r cars}
#Loading data

temp <- tempfile() #Temporary variable to save zip file
url <- "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
download.file(url,temp)
zip_names <- unzip(temp, list = TRUE) #Explore and store zip file
zip_list <- as.vector(zip_names[,1])

```

## Retrieve datasets

The data sets were retrieved using a while loop to go over all the files on the zip folder.

```{r}
#Retrieve and store tables in dataframes

i <- 1
for(i in 1:length(zip_list)) {                    # assign function within loop

  if (zip_names[i,2] != 0) { #Check if the file is empty and skip if it is
    assign(gsub(".*\\/", "", (zip_names[i,1])), read.table(unz(temp, zip_names[i,1]), fill = T))
    i < i + 1
  } else {
    i < i + 1
  }
  
}
```

After extracting the relevant files, I deleted the temp file

```{r}
unlink(temp) #Delete temporary file
```

## Tidying data

After working on this, I added names to each variables within the relevant **x** and **y** tables.

```{r pressure, echo=FALSE}
# Tidy datasets

colnames(X_test.txt) <- features.txt$V2 # Assign correct column names
colnames(X_train.txt) <- features.txt$V2
colnames(activity_labels.txt) <- c("activity", "label")
```

Using the special characters I have extracted the relevant columns (the ones containing the mean values) from the two tables.

```{r}
# Retrieve the mean and std columns

X_train <- X_train.txt[, grep("mean\\(\\)|std\\(\\)", colnames(X_train.txt))]
X_test <- X_test.txt[, grep("mean\\(\\)|std\\(\\)", colnames(X_test.txt))]
```

After this, I  added the activity and subject ids to the table.

```{r}

# Add ID fields

X_test$subjectid <- subject_test.txt$V1
X_train$subjectid <- subject_train.txt$V1

X_test$activity <- y_test.txt$V1
X_train$activity <- y_train.txt$V1

# Merge data sets

X_test <- join(X_test, activity_labels.txt, by = "activity")
X_train <- join(X_train, activity_labels.txt, by = "activity")

```

Once both tables have all relevant values, I merged them to create the summary table.

```{r}
# Create summary

  # Merge both tables

full_data <- rbind(X_train, X_test)

  #Summary
summary_data <- full_data %>% 
  group_by(subjectid, label, activity) %>%
  summarise_all("mean")

```

## Save final dataset

After creating tidy data, we save it under the data directory.

```{r}
# Save file

write.csv(summary_data, file = "./data/mean_values.csv")
```

